\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2014-05-06 rebuttal.tex

\documentclass{article}
\usepackage[utf8]{inputenc}

\begin{document}

\section{General Comments}

    \item \emph{The form and depth of the presentation are below the standards of an
AIJ article.  The reviewers made many suggestions to make the presentation
better structured, more detailed, and more focused.}

@@

    \item \emph{The form and depth of the presentation are below the standards of an
AIJ article.  The reviewers made many suggestions to make the presentation
better structured, more detailed, and more focused.}

We would like first to thank the reviewers for their thoughtful review of our
work. The manuscript underwent a major rewrite, impacting both the overall style
(which should now be much more pleasant to read) and the whole organisation of
the paper (following the reviewers' recommendations). We also dedicated time to
redraw the figures in an improved academic style.

We address hereafter each of the specific concerns raised by the reviewers (note
that minor typos/style issues have been fixed directly in the manuscript, and
are not discussed here).

\item \emph{The novelty of this paper with respect to other papers by the same
authors, including one in IJSR 2011, has been questioned. The authors should
convincingly explain what are the novel contributions and the added value of
this work compared to the previous publications.}

@@

    \item \emph{p 20: AFAIK KnowRob stores OWL as facts in Prolog, why
        should the TBOX be immutable?} The reviewer is right insofar as there is
        not limitations inherent to Prolog in that regard. We have removed the
        reference to KnowRob TBOX, that originated in a discussion with KnowRob's author who told
        us then that KnowRob could not update at run-time its TBox. Since we were
        not able to find litterature confirming this fact, it is likely better
        not to mention it.

\end{itemize}

\section{Comments from Reviewer 2}

Reviewer 2 raised several concerns regarding the overall article organisation
and lack of focus: \emph{``The overall impression is that the paper tries to
cove a too brad set of topics. As a result the takeaway message for the user is
too vague.''}

S/he recommended to modify the section 3 to either provide more informations
\emph{``to understand the problems and the technical solutions''} or to focus on
one particular cognitive skill; asked for \emph{``more substantial validation
of the specific problem(s) chosen in section 3''} in section 4, and suggested
several modifications to the discussion.

We have attempted to address these concerns in the following way:

\begin{itemize}
    \item section 3 underwent a large rewrite, in particular to provide
        technical insights, as well as a better overall picture of the role and
        importance of these skills. We have also added a subsection on the
        reasoning processes that take place in the background.
    \item the section 4 (previously ``experimental outcomes'') has been
        rewritten to make clear that we only present here \emph{case studies}
        that are meant to illustrate and clarify how the whole architecture
        works ``in the real world''. Besides, more details have been provided to the
        \emph{Clean the Table} study (as well as a redesigned figure) to help
        the user to follow the execution flow.
    \item the discussion section has been entirely reorganised, following the
        Reviewer recommendation. The key decisional issues have been explicitly
        stated, and overlaps between subsections have been addressed.
\end{itemize}

We address hereafter the other specific points raised by the reviewer:

    \item \emph{pg. 7 (Section 3.1) ``From an artificial intelligence
            perspective, Spark takes care of merging sensing modalities, it
            computes and grounds perceptions into symbolic knowledge (situation
            assessment), it is responsible temporal interpretation of situations
            and performs spatial perspective-taking.'' given on this
        characterization, Spark should be discussed in greater detail.} We have
        significantly developed the presentation of {\sc Spark}, in
        particular to exhaustively cover what and how symbolic relations are
        computed.

    \item \emph{pg 10 (section 3.2.1)
            "The system supports ...". This is one example of a topic that would
            be really interesting to the readers and it is just hinted. In
            particular, learning rules like "cats are animals" maybe really
            nice, but are such rules compatible with the existing knowledge?
            Adding them as triples may work only in absence of contradictions} The
            various natural language grounding processes that are made possible
            within this semantic
            architecture would require a relatively long discussion to be adequately
            explained. They are besides already presented in~\cite{Lemaignan2011a}.
            We have added an explicit reference to this previous publication.

### 2014-05-07 paper.tex
Proof-reading of sections 1,2,3 and 4

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities, and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
domestic tasks by taking into account, at every stages, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate to communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part to joint actions, both
proactively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

\caption{The robot reasons and acts in domestic interaction scenarios.  The
    sources of information are multi-modal dialogue {\it (A)}, and
    perspective-aware monitoring of the environment and human activity {\it
    (B)}. The robot must adapt on-line its behaviours by merging computed plans
    with reactive control. The robot explicitly reasons on the fact that it is
    (or not) observed by the human. Reasoning and planning take place at
    symbolic {\it (C)} as well as geometric level and take into account agents
    beliefs, perspectives and capabilities {\it (D)} as estimated by the robot.}

@@

\emph{Communication}, \emph{Joint Actions}, \emph{Human-Aware Execution}: these
three challenges structure the whole field of human-robot interaction research,
and can be analysed in terms of the cognitive skills that they require.

@@

The robot controller then decides on an action to execute next and who has to
perform it, from the human, the robot or both in case of a joint action, and
eventually controls or monitors its execution. The operation continues until the
goal is achieved, is declared unachievable or is abandoned by the human.

\begin{inparaenum}
This translates into several decisional, planning, representation skills that
need to be available to the robot. It must be able \item to represent and
manipulate symbolic belief states, \item to acquire and maintain them up-to-date
with respect to state of the world and the task at hand, \item to build and
maintain, and iteratively refine shared (human-robot) plans, \item to
instantiate and execute the actions it has to perform, and conversely,
to monitor those achieved by its human partner.
\end{inparaenum}

Such abilities must be designed and implemented in generic ways and
provide several levels of parametrization such as they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or proactive helper.

@@

The next section presents the overall architecture and the knowledge model we
have developed for our robots. Section~\ref{sec:impl} gives some insights on
each of the main components of the architecture, and try to highlight their
significance for artificial intelligence. Section~\ref{sec:expe} presents two
experiments that illustrate in a concrete way what can be currently achieved
with our robots.  Section~\ref{sect|conclusion} finally summarises our main
contributions and relate them to other recent researches in the field.

@@

\tikzset{subpart/.style={draw, font=\scriptsize, fill opacity=0.5, text opacity=1, fill=black!50}}

@@

\caption{Overview of the architecture. The deliberative layer (on the top),
composed of six main modules, interacts with the low-level sensori-motor
layer (on the bottom). Knowledge is centrally managed in an active
\emph{semantic blackboard}, pictured above with a thick border. The links
between components depicted on the figure underline the central role of the
knowledge base: most of the datastreams are actually symbolic statements exchanged
through the semantic blackboard.}

@@

\subsection{Building a human-aware deliberative layer}

@@

Figure~\ref{fig|archi} gives an overview of our architecture. An active
knowledge base ({\sc Oro}), conveniently thought as a \emph{semantic
blackboard}, connects most of the modules together: the \emph{geometric
reasoning} module (called {\sc Spark}) produces symbolic assertions (like
\stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
environment. These logical statements are stored in the knowledge base, and
reused by the language processing module ({\sc Dialogs}), the symbolic task
planner (HATP) and the execution controller. The output of the language
processing module, or the activities started by the robot controller are
likewise stored as symbolic statements.

For instance, when processing a sentence like ``give me another book'', the {\sc
Dialogs} module sends at some point a request to the knowledge base that looks
like {\tt \footnotesize find(?book type Book, ?book differentFrom
BOOK1)}\footnote{We present a similar but complete example at
section~\ref{moving-london}.} and may write back assertions like \stmt{HUMAN
desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. The HATP
planner then uses the knowledge base to initialise the planning domains with
similar requests ({\tt \footnotesize find(BOOK2 isAt ?location)}, etc.). The
execution controller typically monitor conditions (by subscribing to events
like: {\tt \footnotesize onNewMatch(HUMAN desires ?goal)}) and stores what the
robot is currently doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).

To some extend, this architecture moves away from classical layered approaches
found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
Interactions between components at the deliberative level are mostly
bidirectional and we do not introduce layers of abstraction amongst software
components\footnote{We do have lower-level modules to execute actions or manage
sensors, but all cognition-related modules live in the same global deliberative
layer.}. Dialogue processing, for instance, illustrates this structure. This
component does not purely act as an alternative sensing modality that would be
fed to a distinct deliberative component: on the contrary, it actively queries
the knowledge base to interpret and disambiguate natural language, and in that
regards, it is fully integrated to the deliberative process itself (we detail
natural language processing at section~\ref{sect|com}).

@@

As for any cognitive system, this fundamental interaction between knowledge and
actions is central to our approach, and involves the dialogue component that
acquires \emph{desires} from the other agents, the planner and the execution
controller that may first decide to take or not into account an incoming desire
as a \emph{goal}, and then generates and manages \emph{intentions} from these
goals with the help of a symbolic task planner.

Other activities are however conducted in parallel, without being explicitly
triggered by \emph{desires} in the BDI sense, like situation assessment and
monitoring or other forms of dialogue (including performative dialogue that can
possibly change the internal state of the robot, but does not lead directly to
the creation of \emph{desires}, like assertion of new facts or question
answering).

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the deliberative components (the
clients). It exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to
query the knowledge base. The clients of the {\sc Oro} server are in charge of
managing themselves the knowledge (when to add, when to update, when to
retract knowledge): no meta-semantics is carried over that would let the server
manage these dynamics itself.\footnote{There is one exception: so-called
\emph{memory profiles} let the server discard facts after a specific period
of time. We present this experimental feature at section~\ref{memory}.} We
represent knowledge as RDF triples in the OWL sub-language. Each time
triples are added or removed from the knowledge base, a Description Logics
reasoner ({\sc Pellet}\footnote{\tt http://clarkparsia.com/pellet/})
classifies the whole ontology and inserts all possible inferred triples.

This architecture design (a central knowledge base that essentially appears as a
passive component to the rest of the system -- even though it actually actively
processes the knowledge pool in the background, for instance with the reasoner)
departs from other approaches like the CAST model~\cite{Hawes2007} where
knowledge is represented as a diffuse resource, or the CRAM/KnowRob
architecture~\cite{Beetz2010} where the knowledge base is an active hub that
pro-actively queries perceptual components to acquire knowledge. We believe that
our design leads to a good \emph{observability} (knowledge flows are explicit
and easy to capture since they are centralised) as well as high modularity
(modules communicate through a defined and unified API).

@@

Contrary to other projects like KnowRob~\cite{Tenorth2009a} that relies on
the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
reasoning: at any time, the full belief state is explicitly stated, and
provides the reasoner with the largest possible inference domain. This is of
special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
we are ``looking for'', and we need instead greedily infer as much as possible
to build a belief state as comprehensive as possible.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of important {\sc
OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
robot can be shared or extended with well-defined semantics. A good example is
the concept of \emph{Object}: In everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. Normally, a
human is not considered as an object. In {\sc Cyc}, an object has a more precise
definition: it is something \emph{partially tangible}. That includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange informations between artificial
systems.

@@

\concept{Bottle} $\equiv$ \concept{Container} {\bf and} \concept{Tableware}
{\bf that} (\concept{hasShape} {\bf value} \concept{cylinderShape} {\bf and}
\concept{hasCharacteristicDimension} {\bf only} \concept{\em int[>= 0.1, <=
0.3]})

@@

\emph{Symbol grounding} relates the knowledge model to the perception and
actuation capabilities of the robot, and our architecture adopts here a
bidirectional approach.  The different components that we have described so far
provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
processing modules constantly build and push new symbolic contents about the
world to the knowledge base where it becomes accessible to decisional layers. In
parallel, the knowledge base relies on reasoning in a \textit{top-down} way to
produce new facts, \ie new beliefs that may lead to actual decision making and
physical actions.

@@

While the previous section provided an overview of our cognitive architecture
as well as its associated knowledge model, this section discusses each of its main
``building blocks''. Each of these components may have connections to several
others, and we invite the reader to refer to Figure~\ref{fig|archi} as a guide
if needed.

@@

As mentioned in the previous section, we use the Pellet open-source reasoner to
reason on the knowledge base. This enables several standard inference services:
\emph{consistency checking}, \emph{concept satisfiability},
\emph{classification} and \emph{realisation} (computation of the most specific
classes that an individual belongs to). In case of logical inconsistency, the
reasoner can also provide \emph{explanations} that are important for debugging.

Besides those, {\sc Oro} server implements several algorithms to identify
similarities and differences between concepts (classes or
instances)~\cite{Ros2010b}: the \emph{Common Ancestors} algorithm, useful to
determine the most specific class(es) that include a given set of individuals;
the \emph{First Different Ancestors} algorithm that returns what can be
intuitively understood as the most generic types that \emph{differentiate} two
concepts; and \emph{clarification} and \emph{discrimination} algorithms that
play a key role in the process of \emph{interactive grounding} of the semantics
of concepts (we discuss this process at section~\ref{sect|com}). Clarification
and discrimination algorithms are based on what we call \emph{descriptors}, \ie
properties of individuals, either statically asserted in the common-sense
ontology, acquired by the robot through perception or proactive questioning of
the human partner, or derived from other reasoning algorithms like the
\emph{Common Ancestors} and \emph{Different Ancestors}. The
\emph{discrimination} algorithm consists then in looking for discriminants, \ie
descriptors that allow a maximum discrimination among a set of individuals.

@@

From a robotics point of view, it supposes the ability to build, store and
retrieve separate models of the beliefs of the agents the robot interacts with.
Our knowledge base implements such a mechanism: when the robots recognises that
a new agent has been created in the knowledge base, it initialises a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: the robot (geometrically) computes that the book is in its
own field of view, but not in the human one. The robot knowledge contains the
fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false} (this computation, called \emph{perspective taking}, is
discussed in the next section).

@@

Memory has been studied at length in the cognitive psychology and
neuropsychology communities: very briefly, the idea of \emph{short-term} and
\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968} ,
Anderson~\cite{Anderson1976} splits memory into \emph{declarative} (explicit)
and \emph{procedural} (implicit) memories, Tulving~\cite{Tulving1985} organises
the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
into a hierarchy. Short-term memory is eventually refined with the concept of
\emph{working memory} by Baddeley~\cite{Baddeley2010}

@@

The {\sc Oro} server features a mechanism to mimic only minimalistic forms of memory
families.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.  Three such profiles are
predefined: {\tt short term}, {\tt episodic} and {\tt long term}. They are
currently attached to different lifetime for the statements (respectively 10
seconds, 5 minutes and no time limit). After this duration, the statements are
automatically removed from the knowledge base.

This approach is limited. In particular, \emph{episodic} memory should primarily
refer to the semantics of the statements (that is expected to be related to an
event) and not to a specific life duration.

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\tt short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}: If a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes.

In the larger field of cognitive architectures, the {\sc Soar}
architecture~\cite{Lehman2006} is one of those that tries to reproduce a
human-like memory organisation. The GLAIR cognitive
architecture~\cite{Shapiro2009} also has a concept of long term/short term and
episodic/semantic memories.

@@

Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
grounded amodal representations~\cite{Mavridis2006}, semantic maps
(\cite{Nuechter2008, Galindo2008,Blodow2011}) or affordance-based planning and
object classification~\cite{Lorken2008, Varadarajan2011}.

@@

We rely on a dedicated geometric and temporal reasoning module called {\sc
Spark} (\emph{SPAtial Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a
situation assessment reasoner that generates symbolic knowledge from the
geometry of the environment with respect to relations between objects,
robots and humans (Figures~\ref{fig|spark-oro}
and~\ref{fig:sparkSubfigures}), also taking into account the different
perspective that each agent has on the environment.

{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is
continuously updated at run-time by the robot based on its sensors (in our
experiments, objects are identified and localised through 2D fiducial markers, while
humans are tracked with Kinect-like devices, optionally assisted by motion
capture to accurately track the head motion, which is required to compute what
the human is looking at).

@@

From an artificial intelligence perspective, {\sc Spark} takes care of merging
sensing modalities, it computes and grounds perceptions into symbolic
knowledge (situation assessment), it is responsible for temporal interpretation of
situations and performs spatial perspective-taking.

>>>>

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities, and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
domestic tasks by taking into account, at every stages, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate to communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part to joint actions, both
pro-actively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

\caption{The robot reasons and acts in domestic interaction scenarios.  The
    sources of information are multi-modal dialogue {\it (A)}, and
    perspective-aware monitoring of the environment and human activity {\it
    (B)}. The robot must adapt on-line its behaviours by merging computed plans
    {\it (C)} with reactive control. The robot explicitly reasons on the fact
    that it is (or not) observed by the human. Reasoning and planning take place
    at symbolic as well as geometric level and take into account agents beliefs,
    perspectives and capabilities {\it (D)} as estimated by the robot.}

@@

\emph{Communication}, \emph{Joint Actions}, \emph{Human-Aware Execution}: these
three challenges structure the whole field of human-robot interaction research,
and they can be analysed in terms of the cognitive skills that they require.

@@

The robot controller (with the help of a task planner) decides what action to
execute next, and who should perform it from the human or the robot (or both in
case of a joint action), and finally controls or monitors its execution. The
operation continues until the goal is achieved, is declared unachievable or is
abandoned by the human.

\begin{inparaenum}

This translates into several decisional, planning, representation skills that
need to be available to the robot. It must be able \item to represent and
manipulate symbolic belief states, \item to acquire and maintain them up-to-date
with respect to state of the world and the task at hand, \item to build and
iteratively refine shared (human-robot) plans, \item to instantiate and execute
the actions it has to perform, and conversely, to monitor those achieved by
its human partner.

\end{inparaenum}

Such abilities must be designed and implemented in generic ways and
must provide several levels of parametrization such as they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or pro-active helper.

These are the challenges that we will discuss in this article.

@@

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
section~\ref{sec:impl} gives some insights on each of the main components of the
architecture, and try to highlight their significance for artificial
intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
practical way what can be currently achieved with our robots.
Section~\ref{sect|conclusion} finally restate our main contributions and discuss
the key challenges that human-robot interaction brings to artificial
intelligence.

@@

\tikzset{subpart/.style={draw, font=\scriptsize, fill opacity=0.5, text opacity=1, fill=white!50}}

@@

\caption{Overview of the architecture. A deliberative layer, composed of six
    main modules, interacts with a low-level sensori-motor layer. Knowledge is
    centrally managed in an active \emph{semantic blackboard}, pictured above
    with a thick border. The links between components depicted on the figure
    underline the central role of the knowledge base: most of the datastreams are
    actually symbolic statements exchanged through this semantic blackboard.}

@@

\subsection{Building a Human-Aware Deliberative Layer}

@@

Figure~\ref{fig|archi} gives an overview of our architecture. An active
knowledge base ({\sc Oro}), conveniently thought as a \emph{semantic
blackboard}, connects most of the modules together: the \emph{geometric
reasoning} module ({\sc Spark}) produces symbolic assertions (like
\stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
environment. These logical statements are stored in the knowledge base, and
queried back by the language processing module ({\sc Dialogs}), the symbolic task
planner (HATP) and the execution controller. The output of the language
processing module and the activities started by the robot controller are
likewise stored as symbolic statements.

For instance, when processing a sentence like ``give me another book'', the {\sc
Dialogs} module queries the knowledge base: {\tt \footnotesize find(?book type
Book, ?book differentFrom BOOK1)}\footnote{We present study with a complete
example at section~\ref{moving-london}.}, and write back assertions like
\stmt{HUMAN desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc
Oro}. The HATP planner then uses the knowledge base to initialise the
planning domains with similar requests ({\tt \footnotesize find(BOOK2 isAt
?location)}, etc.), and the execution controller typically monitor
conditions (by subscribing to events like: {\tt \footnotesize
onNewMatch(HUMAN desires ?goal)}) and stores what the robot is currently
doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).

To some extend, this architecture moves away from classical layered approaches
found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
Interactions between components at the deliberative level are mostly
bidirectional and we do not introduce layers of abstraction amongst software
components\footnote{We do have lower-level modules to execute actions or manage
sensors, but all cognition-related modules live in the same global deliberative
layer.}. Dialogue processing, for instance, illustrates this structure. This
component does not purely act as an alternative sensing modality that would be
fed to a distinct deliberative component. On the contrary, it actively queries
the knowledge base to interpret and disambiguate natural language, and in that
regards, it is fully integrated to the deliberative process itself (we detail
natural language processing at section~\ref{sect|com}).

@@

As for any cognitive system, this fundamental interaction between knowledge and
actions is central to our approach as well, and typically involves the dialogue
module to acquire \emph{desires} from the other agents, and the planner and the
execution controller that to first decide to take or not into account an
incoming desire as a \emph{goal}, and then to generate and manage
\emph{intentions} from these goals through the symbolic task planner.

Building upon a BDI architecture, other activities are conducted in parallel,
without being explicitly triggered by \emph{desires} in the BDI sense, like
situation assessment and monitoring or other forms of dialogue (including
performative dialogue that can possibly change the internal state of the robot,
but does not lead directly to the creation of \emph{desires}, like assertion of
new facts or question answering).

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the other deliberative components (the
clients). It exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to
query the knowledge base.  We represent knowledge as RDF triples in the OWL
sub-language. Each time triples are added or removed from the knowledge base, a
Description Logics reasoner ({\sc Pellet}\footnote{\tt
http://clarkparsia.com/pellet/}) classifies the whole ontology and inserts all
possible inferred triples.  The clients of the {\sc Oro} server are in charge of
managing themselves the knowledge (when to add, when to update, when to retract
knowledge): no meta-semantics is carried over that would let the server manage
these dynamics itself.\footnote{There is one exception: so-called \emph{memory
profiles} let the server discard facts after a specific period of time. We
present this experimental feature at section~\ref{memory}.}

This architecture design (a central knowledge base that essentially appears as a
passive component to the rest of the system -- even though it actually actively
processes the knowledge pool in the background, for instance with the reasoner)
departs from other approaches like the CAST model~\cite{Hawes2007} where
knowledge is represented as a diffuse, pervasive resource, or the CRAM/KnowRob
architecture~\cite{Beetz2010} where the knowledge base is an active hub that
pro-actively queries perceptual components to acquire knowledge. We believe that
our design leads to a good \emph{observability} (knowledge flows are explicit
and easy to capture since they are centralised) as well as high modularity
(modules communicate through an explicit and unified API).

@@

Contrary to other projects like KnowRob~\cite{Tenorth2009a} that relies on
the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
reasoning: at any time, the full belief state is explicitly stated, and
provides the reasoner with the largest possible inference domain. This is of
special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
we are ``looking for'', and we need instead to greedily infer as much as possible
to build a belief state as comprehensive as possible.

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
robot can be shared or extended with well-defined semantics. A good example is
the concept of \emph{Object}: In everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. Normally, a
human is not considered as an object. In {\sc Cyc}, an object has a more precise
definition: it is something \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange informations between artificial
systems in a non-ambiguous way.

@@

$$
\text{\tt Bottle} \equiv \text{\tt Container} \wedge 
                         \text{\tt Tableware} \wedge
                         \text{\tt hasShape} \in \text{\tt CylinderShape}^{\mathcal{I}} \wedge
                         \text{\tt hasMainDimension} \in [0.1, 0.3]
$$

@@

\emph{Symbol grounding} connects hence the knowledge model to the perception and
actuation capabilities of the robot, and our architecture adopts here also a
bidirectional approach.  The different components that we have mentioned so far
provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
processing modules constantly build and push new symbolic contents about the
world to the knowledge base where it becomes accessible to decisional layers. In
parallel, the knowledge base relies on reasoning in a \textit{top-down} way to
produce new facts, \ie new beliefs that may lead to actual decision making and
physical actions.

@@

While the previous section provided an overview of our cognitive architecture
as well as its associated knowledge model, this section discusses its main
``building blocks''. Each of these components may have connections to several
others, and we invite the reader to refer to Figure~\ref{fig|archi} as a guide
if needed.

@@

As mentioned in the previous section, we use the Pellet open-source reasoner to
reason on the knowledge base. This enables several standard inference services:
\emph{consistency checking}, \emph{concept satisfiability},
\emph{classification} and \emph{realisation} (computation of the most specific
classes that an individual belongs to). In case of logical inconsistency, the
reasoner can also provide \emph{explanations} that are used in particular for
debugging.

Besides those, {\sc Oro} server implements several algorithms (presented
in~\cite{Ros2010b}) to identify
similarities and differences between concepts (classes or
instances): the \emph{Common Ancestors} algorithm, useful to
determine the most specific class(es) that include a given set of individuals;
the \emph{First Different Ancestors} algorithm that returns what can be
intuitively understood as the most generic types that \emph{differentiate} two
concepts; and \emph{clarification} and \emph{discrimination} algorithms that
play a key role in the process of \emph{interactive grounding} of the semantics
of concepts (we discuss this process at section~\ref{sect|com}). Clarification
and discrimination algorithms are based on what we call \emph{descriptors}, \ie
properties of individuals, either statically asserted in the common-sense
ontology, acquired by the robot through perception or pro-active questioning of
the human partner, or derived from other reasoning algorithms like the
\emph{Common Ancestors} and \emph{Different Ancestors}. The
\emph{discrimination} algorithm consists then in looking for discriminants, \ie
descriptors that allow a maximum discrimination among a set of individuals.

@@

From a robotics point of view, it supposes the ability to build, store and
retrieve separate models of the beliefs of the agents the robot interacts with.
Our knowledge base implements such a mechanism: when the robots recognises that
a new agent has been created in the knowledge base, it initialises a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: if the robot (geometrically) computes that the book is in its
own field of view, but not in the human one, the robot knowledge contains the
fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false} (this computation, called \emph{perspective taking}, is
discussed in the next section).

@@

Memory has been studied at length in the cognitive psychology and
neuro-psychology communities: to give the main pointers, the idea of \emph{short-term} and
\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative} (explicit)
and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985} organises
the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
into a hierarchy. Short-term memory is eventually refined with the concept of
\emph{working memory} by Baddeley~\cite{Baddeley2010}.
In the field of cognitive architectures, the {\sc Soar}
architecture~\cite{Lehman2006} is one of those that tries to reproduce a
human-like memory organisation. The GLAIR cognitive
architecture~\cite{Shapiro2009} also has a concept of long term/short term and
episodic/semantic memories.

@@

The {\sc Oro} server features a mechanism to mimic only minimalistic forms of memory
families.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.  Three such profiles are
predefined: {\it short term}, {\it episodic} and {\it long term}. They are
currently attached to different lifetime for the statements (respectively 10
seconds, 5 minutes and no time limit). After this duration, the statements are
automatically retracted.

This approach is limited. In particular, \emph{episodic} memory should primarily
refer to the semantics of the statements (that is expected to be related to an
event) and not to a specific lifespan.

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\it short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}: If a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes.

@@

Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
grounded amodal representations~\cite{Mavridis2006}, semantic
maps~\cite{Nuechter2008, Galindo2008,Blodow2011} or affordance-based planning
and object classification~\cite{Lorken2008, Varadarajan2011}.

@@

We rely on a dedicated geometric and temporal reasoning module called {\sc
Spark} (\emph{SPAtial Reasoning \& Knowledge}, \cite{Sisbot2011}). It is a
situation assessment reasoner that generates symbolic knowledge from the
geometry of the environment with respect to relations between objects,
robots and humans (Figures~\ref{fig|spark-oro}
and~\ref{fig:sparkSubfigures}), also taking into account the different
perspective that each agent has on the environment.

{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is
updated at run-time by the robot based on its sensors (in most of
our studies, objects are identified and localised through 2D fiducial markers,
while humans are tracked with Kinect-like devices, optionally assisted by motion
capture to accurately track the head motion, which is required to compute what
the human is looking at).

@@

{\sc Spark} is designed to run and continuously update the knowledge base (in
our experiments, it runs at about 10Hz). At each steps, it re-computes spatial
relationships and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit major
performance issues in our constrained scenarios), but greatly simplifies the
management of the \emph{dynamics} of the knowledge (\emph{When do I discard
outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
of the reasoner domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).
